Comparative Evaluation of Spark and Flink Stream Processing

### 2017-03-09 paper/paper.tex
[paper] updated throughput results

\vspace{4pt}
\centering
 Ehab Qadah\\
 Supervisor: PD Dr. Michael Mock.
 \vspace{4pt}

@@

\end{abstract}
Recent years have witnessed advances in the  development of Big Data frameworks like Apache Spark and Apache Flink. For instance, both frameworks support real-time data 
stream processing. The question of which platform is superior to the other still open. 
In this report, we provide a performance comparison of streaming data processing in Apache Spark and Apache Flink by measuring different performance metrics (latency and throughput). In addition, we cover some key aspects of streaming data applications and how they are handled in the two frameworks. The experiment is done using multiple streaming workloads over real-world datasets. Results show that Flink outperform Spark streaming in term of latency. While Spark Streaming gives higher throughput rates than Flink. 
\section{Introduction}

@@

  \centering
    \includegraphics[width=.9\textwidth, height=.2\textheight]{streaming-flow.png}
     \caption{ Process flow of Spark Streaming \cite{spark_streaming}.}
\end{figure}

@@

  \centering
    \includegraphics[width=\textwidth, height=.3\textheight]{flink_engine.png}
     \caption{ An example of data flow graph in Flink \cite{flink}.}
\end{figure} 
\subsection{Apache Kafka}
\par Apache Kafka is scalable, fault-tolerant and distributed streaming framework \cite{kafka}. It allows to publish and subscribe to data streams.
Kafka manages the stream records in different categories (i.e., topics) that are partitioned and distributed over the servers of the Kafka cluster. It allows the data producers to publish a stream of records to one or more  Kafka topic. Additionally, consumer applications can subscribe to one or more topic to read data streams. The consumers within a group use the  same group name, which allows Kafka to distribute and balance the stream partitions among the  members of a certain group for the sake of scalability. Figure 3 shows how 4 partitions of a stream are distributed between two groups of consumers based on the number of members. 
Apache Kafka has been widely  adopted, for example, 
Apache Spark and Apache Flink can ingest Kafka streams.

@@

  \centering
    \includegraphics[width=\textwidth, height=.3\textheight]{kafka_groups.png}
     \caption{ Distribution of a stream partitions for consumer groups in Kafka \cite{kafka}.}
\end{figure}

\section{Experiment Setup and Implementation}
This section describes the data stream setup, the design of evaluation streaming workloads and the  implementation details of the proposed solution of Apache Spark and Apache Flink.

@@

  \centering
    \includegraphics[width=.9\textwidth, height=.2\textheight]{kafka.png}
     \caption{The setuop of the Data Stream Producer \& Kafka Cluster.}
\end{figure}

\subsection{Evaluation Streaming Workloads}
In our experimental evaluation, we developed two real-time stream processing workloads that read a ADS-B messages stream from Kafka in order to perform basic trajectories analysis methods. The workloads are designed to cover general key aspects of real-time stream processing tasks and evaluate the corresponding solutions in Spark and Flink. The following are the aspects of streaming data processing were covered by our workloads's design:

\begin{itemize}
\item Handling parallel input streams (e.g., Kafka Stream).
\item How to aggregate the state of input stream.
\item Manage the order of stream records.
\item How to provide and update global data model in a stream processing task.
\item Evaluate the performance by measuring the latency and throughput.
\end{itemize}
We give the description of workloads, the relation with these aspects and  the implementation details of Spark's and Flink's solutions in Section 3.2.1 \& 3.2.1. Afterward, In Section 4 we discuss the performance evaluation and analysis of stream processing workloads in the two frameworks. 
The source code of the two streaming workloads and the data stream producers is available in a repository on Github\footnote{\url{https://github.com/ehabqadah/Spark_vs_Flink}}.

@@

\par The implementation of trajectory statistics computation in Spark Streaming, first, it reads the Kafka stream using \texttt{KafkaUtils.createDirectStream} that creates \texttt{DStream} that represents the stream of position messages, in order to scale up and have parallel Kafka stream receivers, a multiple instances of \texttt{DStream} must be created and combined together using the \texttt{union} operation. The output stream of the \texttt{union} is further processed using \texttt{mapToPair}, to map each position message to tuple of aircraft ID and position message. Then, the irrelevant tuples are filtered based on the position message type using  a \texttt{filter} transformation. Afterward, the tuples that share the same aircraft ID are groped together to construct the stream of trajectories (tuples of Id and list of positions), by applying the \texttt{groupByKey} operation. Since the processing  model of Spark Streaming is micro-batch based and stateless, \texttt{updateStateByKey} operation must be used to maintain the state of the trajectories stream. In the context of this workload, the state is the aggregated statistics of trajectory, so the statistics computation is preformed within the custom \texttt{updateStateByKey} function that continuously applied for every batch.
 The state update function uses the last position of trajectory from previous batch to aggregate and compute statistics quantities to each trajectory's position in the new data batch, and the new positions list is kept as the new state for the trajectory. Inside the state update function, the new positions must be sorted to preserve the correct order, since the tuples after \texttt{groupByKey} operation are shuffled across the cluster's  nodes and sent to the state update function.

@@

The goal of sector change detection workload is to detect the  enter or leaving of aircraft from one air sector to other, by processing the real-time stream of aircraft positions. Given that the dataset of sectors (i.e., polygons) is available as reference to assign the corresponding sector for certain aircraft's position. In this workload, we test the aspect of providing the  global model of data (sectors dataset) in streaming processing workload. Moreover, the preserving of the stream's records order and consuming of parallel Kafka streams points are also covered.

@@

 We implemented the sector change detection workload in Spark Streaming by constructing the trajectories stream and assigning the sectors for each position of a trajectory, then a change between the assigned sectors of two consecutive aircraft's positions is marked  as  an transition of sector for certain trajectory. First, a multiple instances of \texttt{DStream} are created using \texttt{KafkaUtils.createDirectStream} to read in  parallel from Kafka, after  all input data streams are combined using \texttt{union} function. Second, the records (i.e., ADS-B messages) of the resulted union stream are parsed and mapped to tuples of (aircraft ID, position message) using defined \texttt{mapToPair} transformation. Third, the unrelated  tuples are filtered based on the message type through a \texttt{filter} operation. Fourth, the trajectories stream is constructed by combining all tuples with the same ID using the \texttt{groupByKey} function. Then, to manage and aggregate the state of trajectories stream between the mini-batches a configured \texttt{updateStateByKey} function is applied. The state update function assign the corresponding sector for each new position, the old positions of a certain trajectory from previous batch are discarded, but only the last position of old batch is kept to be used alongside with the new positions. In addition, the new positions are sorted to maintain the correct order.
 The sector data set is provided using the built-in \texttt{Broadcast} feature that distributes a shared value for  all cluster nodes in efficient matter. Finally, the detection of sector change is done by an extra filtering transformation using a \texttt{filter} operator, by checking the sectors of consecutive positions in a trajectory to find the ones with a change in the sector assignment.

@@

 The Flink's implementation of sector change detection workload, uses the \texttt{FlinkKafkaConsumer09} that reads the position messages (i.e., ADS-B) stream from Kafka, which used to construct \texttt{DataStream} instance. The \texttt{DataStream}'s records are mapped to tuples of (aircraft ID and aircraft's position), by a configured \texttt{map} transformation. The irrelevant tuples are filtered based on the message type using a \texttt{filter} operation. Afterward, the tuples that shares a common IDs are grouped together by the \texttt{keyBy} operator. Since the Flink streaming processes the stream records as it come, a defined \texttt{reduce} operator assign corresponding sector of position on new tuple, while the old tuple with same ID is used to retrieve the previous sector to attach it to the new tuple, while the old tuple is discarded. Finally, the tuples (i.e., tuple of (aircraft ID, position message)) with different previous and current sector are filtered using defined \texttt{filter} transformation.

@@

\par This section presents the results of performance evaluation in Spark (2.0.2) and Flink (1.1.3) for the streaming workloads were described in Section 3.2. Specifically, we provide  the measurements of latency and throughput of the benchmarking workloads execution in Spark Streaming and Flink. We carried the experiment by setup the Kafka stream as described in Section 3.1 on local machine\footnote{ The experiments have been carried  on single machine with Ubuntu 16.04 LTS os, Intel® Core™ i5-6200U CPU @ 2.30GHz × 4  processor, and 16 GB of memory.  }, and execute the different workloads on the same machine (i.e., local mode) to obtain the performance results.

\par We consider the time delay between the streaming time of record in Kafka and the 
finish processing time in a workload as the platform's processing latency (measured in milliseconds). On the other hand, the number of processed stream's records (aircrafts position messages) per second within each streaming workload is the throughput measure. Since the Spark Streaming is micro-batch based,  We examine the effect of different batch intervals
of the Spark Streaming on the performance measurements.
\subsection{Latency Results}

Figure 5 shows the results of the latency measurements of the streaming workloads in Spark and Flink. The observed values present the mean value of multiple trails and measurements. Measurements for different batch durations are given in Spark Streaming, we noticed that making the batch time too small does not give best latency results. Results show that Flink outperform Spark streaming in term of processing latency in the two evaluation workloads.

\begin{figure}[h]
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{latency1.png}
  \caption{Latency of Statistics Computation}

\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{latency2.png}
\caption{Latency of Sector Change Detection }

\end{subfigure}
\caption{Latency Measurements}
\label{fig:fig}
\end{figure}
\subsection{Throughput Results}

The  measurements of throughput for the two streaming workloads are give in Figure 6. The observed values reflect the mean of throughput (message/minute) 
 in Spark streaming and Flink for multiple executions. Also for Spark streaming we tested different 
 batch times. Results show that Spark Streaming has higher throughput than Flink, with increasing the batch duration.

\begin{figure}[h]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{throughput1.png}
  \caption{Statistics Computation}

\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{throughput1.png}
\caption{Sector Change Detection }

@@

In this report, we present an experimental evaluation of the stream processing performance and capabilities of Apache Spark and Apache Flink. We carried our experiment over aircraft positions datasets provided by the Datacorn project, we developed two streaming workloads to cover general aspects of real-time stream processing tasks. Additionally, and we measured the latency and throughput of those workloads solution of Spark and Flink. Results show that Flink outperforms Spark Streaming in term of processing latency. In contrast, Spark Streaming has showed better throughput rates than Flink. 
\par To sum up, Flink's processing model is well-suited to the stream processing tasks that require low latency, also there is no need to manage the Stream's state and batch interval as it is in Spark streaming. Moreover, Flink operates over the individual record of the data stream, while Spark Streaming  processes the input data stream as micro-batches. Thus we can state that  Flink is a true stream processing framework. A side remark, we have  noticed that finding online resources related to Spark is easier than to Flink.

@@

\newpage
\begin{thebibliography}{[MT1]}
%
\bibitem[1]{idc} 
Gantz, John, and David Reinsel. "The digital universe in 2020: Big data, bigger digital shadows, and biggest growth in the far east." IDC iView: IDC Analyze the future 2007.2012 (2012): 1-16.

>>>>

\vspace{6pt}
\centering
 Ehab Qadah\\
 Supervisor: PD Dr. Michael Mock.
 \vspace{4pt}

@@

\end{abstract}
Recent years have witnessed advances in the  development of Big Data frameworks like Apache Spark and Apache Flink. For instance, both frameworks support real-time data 
stream processing. The question of which platform is superior to the other still open. 
In this report, we provide a performance comparison of streaming data processing in Apache Spark and Apache Flink by measuring different performance metrics (latency and throughput). Furthermore, we cover some key aspects of streaming data applications and how they are handled in the two frameworks. The experiment is done using multiple streaming workloads over real-world datasets. Results show that Flink outperform Spark streaming in term of latency. While Spark Streaming gives higher throughput rates than Flink. 
\section{Introduction}

@@

  \centering
    \includegraphics[width=.9\textwidth, height=.3\textheight]{streaming-flow.png}
     \caption{ Process flow of Spark Streaming \cite{spark_streaming}.}
\end{figure}

@@

  \centering
    \includegraphics[width=\textwidth, height=.5\textheight]{flink_engine.png}
     \caption{ An example of data flow graph in Flink \cite{flink}.}
\end{figure} 
\newpage
\subsection{Apache Kafka}
\par Apache Kafka is scalable, fault-tolerant and distributed streaming framework \cite{kafka}. It allows to publish and subscribe to data streams.
Kafka manages the stream records in different categories (i.e., topics) that are partitioned and distributed over the servers of the Kafka cluster. It allows the data producers to publish a stream of records to one or more  Kafka topic. Additionally, consumer applications can subscribe to one or more topic to read data streams. The consumers within a group use the  same group name, which allows Kafka to distribute and balance the stream partitions among the  members of a certain group for the sake of scalability. Figure 3 shows how 4 partitions of a stream are distributed between two groups of consumers based on the number of members. 
Apache Kafka has been widely  adopted, for example, 
Apache Spark and Apache Flink can ingest Kafka streams.

@@

  \centering
    \includegraphics[width=\textwidth, height=.4\textheight]{kafka_groups.png}
     \caption{ Distribution of a stream partitions for consumer groups in Kafka \cite{kafka}.}
\end{figure}

\newpage
\section{Experiment Setup and Implementation}
This section describes the data stream setup, the design of evaluation streaming workloads and the  implementation details of the proposed solution of Apache Spark and Apache Flink.

@@

  \centering
    \includegraphics[width=.9\textwidth, height=.3\textheight]{kafka.png}
     \caption{The setuop of the Data Stream Producer \& Kafka Cluster.}
\end{figure}

\newpage
\subsection{Evaluation Streaming Workloads}
In our experimental evaluation, we developed two real-time stream processing workloads that read a ADS-B messages stream from Kafka in order to perform basic trajectories analysis methods. The workloads are designed to cover general key aspects of real-time stream processing tasks and evaluate the corresponding solutions in Spark and Flink. The following are the aspects of streaming data processing were covered by our workloads's design:

\begin{itemize}
\item Handling parallel input streams (e.g., Kafka Stream).
\item How to aggregate the state of input stream.
\item Manage the order of stream records.
\item How to provide and update global data model in a stream processing task.
\item Evaluate the performance by measuring the latency and throughput.
\end{itemize}
We give the description of workloads, the relation with these aspects and  the implementation details of Spark's and Flink's solutions in Sections 3.2.1 \& 3.2.1. Afterward, In Section 4 we discuss the performance evaluation and analysis of stream processing workloads in the two frameworks. 
\par The two streaming workloads and the data stream producers are implemented in Java and the source code is available in a repository on Github\footnote{\url{https://github.com/ehabqadah/Spark_vs_Flink}}.

@@

\par The implementation of trajectory statistics computation in Spark Streaming, first, it reads the Kafka stream using \texttt{KafkaUtils.createDirectStream} that creates \texttt{DStream} that represents the stream of position messages, in order to scale up and have parallel Kafka stream receivers, a multiple instances of \texttt{DStream} must be created and combined together using the \texttt{union} operation. The output stream of the \texttt{union} is further processed using \texttt{mapToPair}, to map each position message to tuple of aircraft ID and position message. Then, the irrelevant tuples are filtered based on the position message type using  a \texttt{filter} transformation. Afterward, the tuples that share the same aircraft ID are groped together to construct the stream of trajectories (tuples of Id and list of positions), by applying the \texttt{groupByKey} operation. Since the processing  model of Spark Streaming is micro-batch based and stateless, \texttt{updateStateByKey} operation must be used to maintain the state of the trajectories stream. In the context of this workload, the state is the aggregated statistics of trajectory, so the statistics computation is preformed within the custom \texttt{updateStateByKey} function that is continuously applied for every batch.
 The state update function uses the last position of trajectory from previous batch to aggregate and compute statistics quantities to each trajectory's position in the new data batch, and the new positions list is kept as the new state for the trajectory. Inside the state update function, the new positions must be sorted to preserve the correct order, since the tuples after \texttt{groupByKey} operation are shuffled across the cluster's  nodes and sent to the state update function.

@@

The goal of sector change detection workload is to detect the  entering or leaving of an aircraft from one air sector to another one, by processing the real-time stream of aircraft positions. Given the fact that a dataset of sectors (i.e., polygons) is available as a reference to assign the corresponding sector for certain aircraft's position. In this workload, we test the aspect of providing a global data model (sectors dataset) in stream processing workload. Moreover, the preserving of the stream's records order and consuming of parallel Kafka streams  are covered.

@@

 We implemented the sector change detection workload in Spark Streaming by constructing the stream of trajectories and assigning a sector for each position of a trajectory. A change between the assigned sectors of two consecutive aircraft's positions is marked  as a transition of sector for a certain trajectory. \par First, multiple instances of \texttt{DStream} are created using \texttt{KafkaUtils.create} \texttt{DirectStream}  are combined using \texttt{union} to read in  parallel from Kafka. Second, the records (i.e., ADS-B messages) of the resulting stream are parsed and mapped to tuples of (aircraft ID, position message) using \texttt{mapToPair} transformation. Third, the unrelated  tuples are filtered based on the message type using a \texttt{filter} operation. Fourth, the stream of trajectories is constructed by combining all tuples with the same ID using the \texttt{groupByKey} function. Then, to manage and aggregate the state of trajectories stream between the mini-batches a configured \texttt{updateStateByKey} function is applied. The state update function assigns the corresponding sector for each new position, the old positions of a certain trajectory from the previous batch are discarded, but only the last position of the old batch is kept to be used alongside with the new positions. In addition to that the new positions are sorted by th streaming time to maintain the correct order.
 \par The sector data set is provided using the built-in \texttt{Broadcast} feature that distributes a shared value for  all cluster nodes in an efficient matter. Finally, the detection of a sector change is done by an extra filtering transformation using a \texttt{filter} operator, which checks the sectors of consecutive positions in a trajectory to find the ones with a change in the sector assignment.

@@

 The Flink's implementation of sector change detection workload uses the \texttt{FlinkKafkaConsumer09} that reads the position messages (i.e., ADS-B) stream from Kafka, which is used to construct a \texttt{DataStream} instance. The \texttt{DataStream}'s records are mapped to tuples of (aircraft ID and aircraft's position) by a configured \texttt{map} transformation. Irrelevant tuples are filtered based on the message type using a \texttt{filter} operation. Afterward, the tuples that shares a common IDs are grouped together by the \texttt{keyBy} operator. Since the Flink streaming processes the stream records as it come, a defined \texttt{reduce} operator assign corresponding sector of position on new tuple, while the old tuple with same ID is used to retrieve the previous sector to attach it to the new tuple, while the old tuple is discarded. Finally, the tuples (i.e., tuple of (aircraft ID, position message)) with different previous and current sector are filtered using defined \texttt{filter} transformation.

@@

\par This section presents the results of performance evaluation in Spark (2.0.2) and Flink (1.1.3) for the streaming workloads that were described in Section 3.2. Specifically, we provide  the measurements of latency and throughput of the benchmarking workloads execution in Spark Streaming and Flink. We carried the experiment by setting up the Kafka stream as described in Section 3.1 on a local machine\footnote{ The experiments have been carried  on single machine with Ubuntu 16.04 LTS os, Intel® Core™ i5-6200U CPU @ 2.30GHz × 4  processor, and 16 GB of memory.} and execute the different workloads on the same machine (i.e., local mode) to obtain the performance results.

\par We consider the time delay between the streaming time of a record in Kafka and  the time when the record is processed by a workload as the platform's processing latency (measured in milliseconds). And the number of processed stream's records (aircrafts position messages) per second within each streaming workload is the throughput measure. Since the Spark Streaming is micro-batch based,  we examine the effect of different batch intervals
of the Spark Streaming on the performance measurements.

\subsection{Latency Results}

Figure 5 shows the results of the latency measurements of the streaming workloads in Spark and Flink. The presented values present the average value of multiple trails and measurements. Measurements for different batch durations are given in Spark Streaming, we noticed that making the batch time too small does not give better latency results. Results show that Flink outperform Spark streaming in term of processing latency in the two evaluation workloads.

\begin{figure}[h]
\centering
\begin{subfigure}{.7\textwidth}
  \centering
  \includegraphics[width=\linewidth]{latency1.png}
  \caption{Latency of Statistics Computation}

\end{subfigure}%
\vskip\baselineskip
\centering
\begin{subfigure}{.7\textwidth}
  \centering
  \includegraphics[width=\linewidth]{latency2.png}
\caption{Latency of Sector Change Detection }

\end{subfigure}
\caption{Latency Measurements}
\label{fig:fig}
\end{figure}
\newpage
\subsection{Throughput Results}

The  measurements of throughput for the two streaming workloads are given in Figure 6. The presented values reflect the average of throughput (message/minute) 
 in Spark streaming and Flink for multiple executions. Also for Spark streaming we tested different 
 batch times. Results show that Spark Streaming has higher throughput than Flink, also  increasing the batch duration of Spark Streaming gives better throughput.

\begin{figure}[h]
\centering
\begin{subfigure}{.7\textwidth}
  \centering
  \includegraphics[width=\linewidth]{throughput1.png}
  \caption{Statistics Computation}

\end{subfigure}%

\vskip\baselineskip
\centering
\begin{subfigure}{.7\textwidth,}
  \centering
  \includegraphics[width=\linewidth]{throughput2.png}
\caption{Sector Change Detection }

@@

In this report, we present an experimental evaluation of the stream processing performance and capabilities of Apache Spark and Apache Flink. We carried our experiment over aircraft positions datasets provided by the Datacorn project, we developed two streaming workloads to cover general aspects of real-time stream processing tasks. Additionally, and we measured the latency and throughput of those workloads solution of Spark and Flink. Results show that Flink outperforms Spark Streaming in term of processing latency. In contrast, Spark Streaming has showed better throughput rates than Flink. 
\par To sum up, Flink's processing model is well-suited to the stream processing tasks that require low latency, also there is no need to manage the Stream's state and the batch interval in contrast to Spark streaming. Moreover, Flink operates over the individual record of the data stream, while Spark Streaming  processes the input data stream as micro-batches. Thus we can state that  Flink is a true stream processing framework. A side remark, we have  noticed that finding online resources related to Spark is easier than to Flink.

@@

\begin{thebibliography}{[MT1]}
%
\bibitem[1]{idc} 
Gantz, John, and David Reinsel. "The digital universe in 2020: Big data, bigger digital shadows, and biggest growth in the far east." IDC iView: IDC Analyze the future 2007.2012 (2012): 1-16.
